Binning Strategies and Related Loss for Large Data

### 2015-03-05 Paper.Rnw
continued work on frequency binning section.  Still incomplete, needs visual display.

\ktm{ Bin counts can be displayed using a continuous shading scale in HSV color space (reference needed) and thus have no frequency loss. While the tiles of the binned scatterplot would be rendered precisely, the ability of a human with average vision to extract that information using a plot index for the frequency key for the plot is largely imprecise.  Color perception of shade is extremely context sensative (reference here) allowing an inaccurate mapping of tile shade to the corresponding shade on the frequency key.  }

\ktm{The goal of binning the frequencies and using a ordinally shaded color palette is to provide more distinctive perceivable breaks to aid the information extraction; however it does so at the cost of using shade to display binned freqencies, $b_c(c_\ell)$ instead of the true frequencies, $c_\ell$. The \textit{total frequency loss}, $L^F$ are defined as}
%
\begin{eqnarray}\label{totfreqloss}
L^F = \sum_{i=1}^{n} L_i^F = \sum_{i=1}^{n} (c_\ell-b_c(c_\ell))^2
\end{eqnarray}  
%
\ktm{where $L_\ell^F$ is the frequency loss for the $\ell^{th}$ bin. Note that this is effectively a sum of squared error from binning frequencies.}

@@

Frequency data consists of counts, which most commonly exhibit skew densities, i.e. there are usually a lot of cells with small cell counts and a few cells with extremely large counts.

A log transformation therefore produces a more symmetric distribution of frequency information, increasing perceptual resolution. This is consistent with the Weber-Fechner law which suggests that increased stimulus intensity is perceptually mapped on the log scale \citep{sp}. Using a logarithmic mapping of frequency to the color or size aesthetic provides a more natural perceptual experience and simultaneously increases the perceptual resolution of the graph. This suggests that frequency loss be calculated as 
\begin{eqnarray}\label{eqn:logfrequency}
L_{\log \text{Freq}} = \sum_{j=1}^{N_x N_y} \left(\log(c_j+1) - \log(b(c_j)+1)\right)^2
\end{eqnarray}

Note that we have added one count to every cell before taking the logarithm to avoid problems with empty cells.

>>>>

\ktm{ Bin counts can be displayed using a continuous shading scale in HSV color space (reference needed) and thus have no frequency loss. While the tiles of the binned scatterplot would be rendered precisely, the ability of a human with average vision to extract that information using a plot index for the frequency key for the plot is largely imprecise.  Color perception of shade is extremely context sensitive (reference here) allowing an inaccurate mapping of tile shade to the corresponding shade on the frequency key.  }

\ktm{The goal of binning the frequencies and using a ordinally shaded color palette is to provide more distinctive perceivable breaks to aid the information extraction; however it does so at the cost of using shade to display binned frequencies, $b_c(c_\ell)$ instead of the true frequencies, $c_\ell$. The \textit{total frequency loss}, $L^F$, is defined as}
%
\begin{eqnarray}\label{totfreqloss}
L^F = \sum_{\ell=1}^{L} L_\ell^F = \sum_{\ell=1}^{L} (c_\ell-b_c(c_\ell))^2
\end{eqnarray}  
%
\ktm{where $L_\ell^F$ is the frequency loss for the $\ell^{th}$ bin. Note that this is effectively a sum of squared error from binning frequencies.}

Frequency data consists of counts, which most commonly exhibit skew densities, i.e. there are usually a lot of cells with small cell counts and a few cells with extremely large counts. A log transformation therefore produces a more symmetric distribution of frequency information, increasing perceptual resolution. This is consistent with the Weber-Fechner law which suggests that increased stimulus intensity is perceptually mapped on the log scale \citep{sp}. Using a logarithmic mapping of frequency to the color or size aesthetic provides a more natural perceptual experience and simultaneously increases the perceptual resolution of the graph. The \textit{log frequency loss}, $L^{\log F}$, is defined as
%
\begin{eqnarray}\label{logfrequency}
L^{\log F} = \sum_{\ell=1}^{L} L_\ell^{\log F} = \sum_{\ell=1}^{L} (\log(c_\ell+1) - \log(b_c(c_\ell+1)))^2
\end{eqnarray}
%
Note that we have added one count to every cell before taking the logarithm to avoid problems with empty cells.

\ktm{Plot needed here to demonstrate shading scales under binning.}

@@

Frequency data consists of counts, which most commonly exhibit skew densities, i.e. there are usually a lot of cells with small cell counts and a few cells with extremely large counts. A log transformation therefore produces a more symmetric distribution of frequency information, increasing perceptual resolution. This is consistent with the Weber-Fechner law which suggests that increased stimulus intensity is perceptually mapped on the log scale \citep{sp}. Using a logarithmic mapping of frequency to the color or size aesthetic provides a more natural perceptual experience and simultaneously increases the perceptual resolution of the graph. This suggests that frequency loss be calculated as 
%
\begin{eqnarray}\label{eqn:logfrequency}
L_{\log \text{Freq}} = \sum_{j=1}^{N_x N_y} \left(\log(c_j+1) - \log(b(c_j)+1)\right)^2
\end{eqnarray}
%
Note that we have added one count to every cell before taking the logarithm to avoid problems with empty cells.
